---
title: "ggplot2 - January 2019 Data Carpentry Lesson"
author: "Maria Kamenetsky"
date: "January 17, 2019"
output:
  html_document:
    df_print: paged
  pdf_document: default
minutes: 60
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.height = 3, fig.width = 5)
solution <- FALSE
```

#Outline

- 0) Prep the Data
- 1) ```ggplot()``` function
    - base plot
    - x and y
- 2) *geoms*
    - + vs. %>%
- 3) Other Aesthetics
- 4) Layering (line plot + boxplot)
- 5) Facetting
    - ```facet_grid()```
    - ```facet_wrap()```
- 6) Themes
- 7) Saving plots (if time allows)

#  Why ```ggplot2```?

- exploratory data analysis
    - quick exploration for trends in data, potential outliers, etc.
- if you can think of a visualization, there's probably a way to do it using ggplot
- customization is easy
- 1 viz package to rule them all
- syntax is consistent
- **gg** = grammar of graphics. There is a syntax to creating plots, which involves adding things in brick by brick


# 0) Prep the Data

Load packages we will be using:

```{r}
library(dplyr)
library(ggplot2)
```

**GOAL:** Use ```dplyr``` tools to reformat our data so that we can make visualizations for *mammals* data.

- We will make three datasets from *portal_clean.csv*:
    - 1) **just_dm** dataset only containing species=="DM"
    - 2) **stat_summary** dataset containing mean weight, mean height, and observation count by species
    - 3) **year_summary** dataset containing mean weight, mean height, and observation count by species, year, and sex.

Load data:

```{r}
download.file("http://kbroman.org/datacarp/portal_clean.csv",
              "C:/Users/Maria/Desktop/DataCarpentry/Clean/portal_clean.csv")
surveys <- read.csv("C:/Users/Maria/Desktop/DataCarpentry/Clean/portal_clean.csv")
```

Create three datasets:

```{r}
#just_dm
just_dm <- surveys %>% filter(species_id=="DM")
str(just_dm)

#stat_summary
stat_summary <- surveys %>%
    group_by(species_id) %>%
    summarize(mean_wt=mean(weight),
              mean_hfl=mean(hindfoot_length),
              n=n())
str(stat_summary)

#year_summary
year_summary <- surveys %>%
    group_by(species_id, year, sex) %>%
    summarize(mean_wt=mean(weight),
              mean_hfl=mean(hindfoot_length),
              n=n())
str(year_summary)
```



# 1) ```ggplot()``` function

**Goal:** scatterplot of weight (x) by hindfoot_length (y) using *surveys* dataset.


```{r}
ggplot(surveys, aes(x = weight, y = hindfoot_length)) 
```

Empty plot! We need to tell ```ggplot()``` what kind of plot we want. Default is to only plot the axes. To select the plot type, we need to learn about **geom**'s or **geometries**.

# 2) *geom*'s

- *geom* is short for geometry. This determines the type of plot that ggplot will make
- *many* different geom's available
- scatterplot: ```geom_point()```


```{r}
ggplot(surveys, aes(x = weight, y = hindfoot_length)) + geom_point()

```

Can assign this plot to an object:

```{r}
p1 <- ggplot(surveys, aes(x = weight, y = hindfoot_length)) + geom_point()
#nothing happens
```

```{r}
p1
```

This makes it easy to try different things using **+** operator.


```{r}
#log scale for x-axis
p1 + scale_x_log10()
```

```{r}
#square root scale for x-axis
p1 + scale_x_sqrt()
```

#### CHALLENGE 1: Make a scatterplot of *hindfoot_length* vs. *weight* but only for *species_id* "DM"**

- Use the dataset we created,**just_dm**
- Use our ```ggplot2()``` code above but with this new dataset in place of *surveys*.

```{r, include=solution}
#Challenge solution

ggplot(just_dm, aes(x=weight, y= hindfoot_length)) + geom_point()
```

# Other Aesthetics

- shape

```{r}
ggplot(surveys, aes(x = weight, y = hindfoot_length)) +
    geom_point(shape="triangle")
```

```{r}
#assign base plot to p2 to avoid extra typing
p2 <- ggplot(surveys, aes(x = weight, y = hindfoot_length))
```


- point size

```{r}
#These two plots are equivalent because I assigned p2 <- ggplot(surveys, aes(x = weight, y = hindfoot_length))
p2 + geom_point(size=0.5)
# ggplot(surveys, aes(x = weight, y = hindfoot_length)) +
#     geom_point(size=0.5)
```


- color

```{r}
p2 + geom_point(color = "slateblue")
# ggplot(surveys, aes(x = weight, y = hindfoot_length)) +
#     geom_point(color = "slateblue")
```


- alpha (= transparency)


```{r}
p2 + geom_point(alpha = 0.1)
# ggplot(surveys, aes(x = weight, y = hindfoot_length)) +
#     geom_point(alpha = 0.1)
```

- Putting it all together

```{r}
p2 + geom_point(shape="triangle", size=0.5, color="slateblue", alpha=0.1)

```

## Assign Aesthetics to Data

- we can also assign aesthetics to data instead of fixed values or colors.

```{r}
p2 + geom_point(aes(color= species_id))
```

- **CHALLENGE 2:  Make a scatterplot of mean *hindfoot_length* vs. mean *weight*, where each point is a species, and where the sizes of the points indicate the sample size**
    - Use the dataset ```stat_summary```
    - Use our ggplot code with the aesthetics ```x=mean_wt``` and ```y=mean_hfl```, plus ```size=n```.
    
    
```{r, include=solution}
ggplot(stat_summary, aes(x=mean_wt, y=mean_hfl)) + geom_point(aes(color = species_id, size=n))
```
    
# Layering

- recall, grammar of graphics, add things on layer by layer, brick by brick.

##Boxplots

**Goal:** Boxplots of weight by species.

We will use the ```geom_boxplot()``` geom.

```{r}
ggplot(surveys, aes(x=species_id, y = weight)) + 
    geom_boxplot()
```

Can add points to boxplots to get better idea of number of measurements and their distribution. We will add ```geom_jitter()```.

- ```geom_boxplot() + geom_jitter()```

```{r}
ggplot(surveys, aes(x=species_id, y = weight)) + 
    geom_boxplot() +
    geom_jitter(alpha =0.3, color="tomato")
    
```

- ```geom_jitter() + geom_boxplot()```

```{r}
ggplot(surveys, aes(x=species_id, y = weight)) + 
    geom_jitter(alpha =0.3, color="tomato") +
    geom_boxplot() 
    
```

##Line Plots (Time-Series)

**Goal:** Make a lineplot counts of animals by year. We will use ```geom_line()``` geom.

First, create new dataset using ```dplyr```

```{r}
count_by_year <- surveys %>%
    group_by(year) %>%
    tally
str(count_by_year)
```

Next, use ```geom_line()``` geom.

```{r}
ggplot(count_by_year, aes(x=year, y=n)) +
    geom_line()
```

```{r}
ggplot(count_by_year, aes(x=year, y=n)) +
    geom_line() +
    geom_point()
```


We know that since ```geom_point()``` was called after ```geom_line()```, the points are placed on top of the lines. We can confirm this if we plot the lines and points in contrasting colors:


```{r}
ggplot(count_by_year, aes(x=year, y=n)) +
    geom_line(color="lightblue") +
    geom_point(color="violetred")
```


We could also control the aesthetics separately for each layer if we want to:

```{r}
ggplot(count_by_year, aes(x=year, y=n)) + 
    geom_line() +
    geom_point(aes(color=year))

```


Or if we want to color both points and lines:

```{r}
ggplot(count_by_year, aes(x=year, y=n, color=year)) + 
    geom_line() +
    geom_point()

#this is equivalent to:

ggplot(count_by_year, aes(x=year, y=n)) +
    geom_line() +
    geom_point() +
    aes(color=year)

```

- **Challenge 3: **
    - **Use the ```year_summary``` dataset to make a line plot of counts of each species by year, with a different colored line for each species**
    - **Use ```aes(linetype=sex)``` to have different line types for the two sexes**
    
```{r lineplots, fig.height=3, fig.width=5, fig.cap="Time series of species count by sex.", include=solution}
ggplot(year_summary, aes(x=year, y=n)) +
    geom_line(aes(color=species_id, linetype=sex))
```

# Facetting

**General Goal:** Plot data in ```year_summary``` into multiple panels.

## ```facet_wrap()```

- ```facet_wrap()``` geometry extracts plots into an arbitrary number of dimensions to allow the to fit on one page

**Specific Goal:** Plot count by year, with separate lines for sex and separate panels for species.

```{r}
ggplot(year_summary, aes(x = year, y = n)) +
    geom_line(aes(color=sex)) +
    facet_wrap(~ species_id)

```



## ```facet_grid()```


**Specific Goal:** Plot count by year, with separate lines for sex and separate panels for species.


- ```facet_grid()``` geometry allows you to explicitly specify how you want your plots to be arranged via formula notation (```rows ~ columns```) 
    - a ```.``` can be used as a placeholder that indicates only 1 row or 1 column.



- vertical split

```{r}
ggplot(year_summary, aes(x = year, y = n)) +
    geom_line(aes(color=sex)) +
    facet_grid(species_id~.)
```

- horizontal split

```{r}
ggplot(year_summary, aes(x = year, y = n)) +
    geom_line(aes(color=sex)) +
    facet_grid(~species_id)
```



- separate panel for each sex and species


```{r}
ggplot(year_summary, aes(x = year, y = n)) +
    geom_line(aes(color=sex)) +
    facet_grid(species_id ~ sex)
```

- **CHALLENGE 4:**
    - **Use the ```year_summary``` dataset and make scatterplots of mean hindfoot length vs. mean weight (with each point being a species), facetting by year.**
        - **Use aesthetics ```x=mean_wt``` and ```y=mean_hfl```**
        - **Use geom_point(aes(color=species_id, shape=sex))**
        - **Use ```facet_wrap(~year)```**



```{r, include=solution}
ggplot(year_summary, aes(x=mean_wt, y=mean_hfl)) +
    geom_point(aes(color=species_id, shape=sex)) +
    facet_wrap(~year)
```


#Themes

- the grey background is contested. You can apply a variety of themes to the overall appearance of the plot.

For example:

```{r}
#surveys %>% filter(species_id %in% c("DM", "DS", "DO")) %>%
ggplot(surveys,aes(x=weight, y=hindfoot_length)) +
    geom_point(aes(color=species_id)) +
    theme_bw()

```

# Saving Plots

- use the ```ggsave()``` function
- default is to save the last plot you created.
    - good practice to save plot as an object and pass that to ```ggsave()```

```{r}
p <- ggplot(surveys, aes(x=weight, y=hindfoot_length)) + geom_point()
ggsave("scatter.png", p, height=6, width=8)
```

To create a different filetype, just change the extension. FOr example, to make a pdf:

```{r}
ggsave("scatter.pdf", p, height=6, width=8)
```






















