---
title: "Capstone project"
author: "Karl Broman (modified by M.Kamenetsky)"
date: "January 17, 2019"
output: html_document
---

Capstone project for Data Carpentry course.

## 1: Load the data

```{r load_data, message=FALSE}
library(dplyr)
library(ggplot2)
surveys <- read.csv("C:/Users/Maria/Desktop/DataCarpentry/Clean/portal_clean.csv")
```

There are `r ncol(surveys)` columns and `r nrow(surveys)` rows.

## 2: Boxplots of weight by sex

Boxplots of weight by sex.

```{r boxplots_weight_noomit}
surveys %>% select(weight, sex) %>%
    filter(sex != "", !is.na(sex)) %>%
    ggplot() + geom_boxplot(aes(x=sex, y=weight))
```


Boxplots of weight by sex, omitting individuals with missing sex.

```{r boxplots_weight_omit}
surveys %>% select(weight, sex) %>%
    filter(sex != "", !is.na(sex)) %>%
    ggplot() + geom_boxplot(aes(x=sex, y=weight))
```

## 3. Histogram of hindfoot lengths

Histogram of hindfoot length:

```{r hist_hindfootlength_noomit}
surveys %>% filter(!is.na(hindfoot_length)) %>%
    ggplot() + geom_histogram(aes(x=hindfoot_length), bins=150)
```

Histogram of hindfoot length, omitting individuals with missing hindfoot_length:

```{r hist_hindfootlength_omit}
surveys %>% filter(!is.na(hindfoot_length)) %>%
    ggplot() + geom_histogram(aes(x=hindfoot_length), bins=150)
```

## 4. Scatterplots of hindfoot length vs weight for 3 species

Using `dplyr` and `ggplot` together:

```{r scatterplots1}
surveys %>% filter(species_id %in% c("DM", "DO", "DS")) %>%
    filter(!is.na(weight), !is.na(hindfoot_length)) %>%
    ggplot(aes(x=hindfoot_length, y=weight)) +
    geom_point(aes(color=species_id)) + facet_grid(species_id ~ .)
```

Separating out `dplyr` and `ggplot` commands:

```{r scatterplots}
species_subset <- surveys %>% filter(species_id %in% c("DM", "DO", "DS")) %>%
    filter(!is.na(weight), !is.na(hindfoot_length))
    
ggplot(species_subset, aes(x=hindfoot_length, y=weight)) +
    geom_point(aes(color=species_id)) + facet_grid(species_id ~ .)
```



## 5. Plot of counts of "DM" in "Rodent Exclosure" plots over time

```{r plot_counts_by_year}
counts <- surveys %>% filter(species_id == "DM") %>%
    filter(plot_type=="Rodent Exclosure") %>%
    group_by(year) %>%
    tally()

counts %>% ggplot(aes(x=year, y=n)) + geom_line()
```


Alternative:

```{r plot_counts_by_year_al}
counts <- surveys %>% filter(species_id == "DM") %>%
    filter(plot_type=="Rodent Exclosure") %>%
    group_by(year) %>%
    tally()

ggplot(counts,aes(x=year, y=n)) + geom_line()


```

## 6. Table with counts of "DM" by plot_type in 1977

Using base R for table:

```{r table_plot_type}
counts1977 <- surveys %>% filter(species_id == "DM", year==1977) %>%
    group_by(plot_type) %>%
    tally() %>%
    select(plot_type, n)
counts1977
```



Using `kable`:

```{r table_plot_typealt}
counts1977 <- surveys %>% filter(species_id == "DM", year==1977) %>%
    group_by(plot_type) %>%
    tally() %>%
    select(plot_type, n)
knitr::kable(counts1977)
```
