---
title: "RMarkdown - Challenge Examples"
author: "Maria Kamenetsky"
date: "January 6, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Challenge 1

*RMarkdown* is really cool and can be used to make **reproducible** reports.

- Some benefits of using RMarkdown are:
    - combines text, code, figures, tables
    - write papers without having to remember to swap figures, change cell values in table.
    - automatically generate reports
    - *reproducible* documents
    
#Challenge 2

```{r}
library(ggplot2)
surveys <- read.csv("C:/Users/Maria/Desktop/DataCarpentry/Clean/portal_clean.csv")
ggplot(surveys, aes(x=sex, y=weight)) + 
    geom_jitter(aes(alpha=0.3, color=sex)) +
    geom_boxplot(fill=c("red", "blue"))

```


#Challenge 3



```{r, eval=FALSE}
# ```{r boxplotsbyweight, echo=FALSE, fig.height=3, fig.width=5}
# ggplot(surveys, aes(x=sex, y=weight)) + 
#     geom_jitter(aes(alpha=0.3, color=sex)) +
#     geom_boxplot(fill=c("red", "blue"))
```
















